
\begin{acknowledgments}

I feel privileged to have been surrounded by very kind, supportive, and joyful people during the course of my PhD. They have helped me grow not just as a researcher, but as a person. Here, I want to express my heart-felt gratitude to everyone without whom this dissertation would not be possible.

To my advisor, Yanni, your confidence in my work has meant more to me than I can put in words here. Thank you for being the mentor, researcher, collaborator, and writer I aspire to be one day. You have constantly guided me, challenged me, and inspired me in critical and constructive ways. It has been an absolute honor being advised by you. 

To my committee members, Richmond, Noura, Lauren, and Chris, thank you for your support these past few years. You have been very generous with your time and have helped me develop my ideas and writings that have eventually shaped this work in fundamental ways. 

To everyone who funded me in this journey; Carol, thank you for inviting me to join the Women, Science, and Technology Learning Community, and financially supporting the entirety of my PhD, never wavering. You have been a constant calming presence in this journey that was sometimes very messy and overwhelming. And Allen, who funded my summers and supported my conference travels, thank you so much. 

To my internship mentors, Sonam, Juanis, and Sara, thank you for keeping me in touch with the world outside of academia. You all have been invaluable in helping me discover what I truly want out of my PhD. You all continue to be my mentors to this day and for that, I am so grateful. 

To my small but mighty Digital Media family, a lot of us started our grad school during COVID, unaware of what a `normal' PhD life looks like. We together created a new normal for us, a normal that I have greatly enjoyed and that I will miss so much. Pooja, Mohsin, and Sylvia thank you for constantly brainstorming with me, reviewing my writing, cheering me on, pushing me to reach for things I thought I could not achieve, and most importantly for being my friend. It would have been a very lonely journey if it weren't for you all. 

To all other friends within and beyond Georgia Tech, Kiran, Jaya, Nikhil, Sushant, Aditya, Vishal, Manoj, Atefeh, Vaibhav, and so so many more people who I may not be able to name here, thank you for listening to my many many rants, for  helping me disconnect from work when needed, and for making sure I enjoy this journey. 

To my fiancé, Nirbhay, you challenged me to start this journey and held my hand as I completed it. Intellectually, you have been my sounding board, my fresh eyes, who has helped me develop each and every argument I make in this work. Emotionally, you have been my rock, who has kept me in touch with who I want to be and what I want to do in this fleeting life. This journey wouldn't be nearly as exciting or gratifying if it weren't for you.  

And lastly, my family, my parents Prabha and Puneet, my brother Manush, and my sister (in law) Priyanshi; Mom and Dad, the list of things I should thank you for is endless. You have shown me what it means to live a wholesome life, a happy life, a life full of love and laughter. It is a privilege to be your daughter. I owe it all to you both. Manush, my big brother, whose footsteps I first followed but who helped me become my own person in time. Thank you for being there for me in times that truly mattered. And Priyanshi, my new sister, thank you for joining our family and ending my PhD journey on a high! 

Mentors, friends, and family, I will cherish my memories with you all forever. Thank you for everything! 


\end{acknowledgments}
